---
title: "Assignment Two - Linear Regression"
author: "Emilia Carson"
date: "2022-11-14"
output: 
  html_document:
    theme: flatly
    highlight: haddock
    toc: true
    toc_depth: 2
    number_section: false
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Assignment 2: Linear Regression

## Reading in the Data

### Relevant Libraries and Packages

```{r, warning=FALSE, message =FALSE}
library(tidyverse)
library(dplyr)
library(ggplot2) 
library(GGally)            
library(purrr)
```

#### Step 1: Reading in the data

```{r, warning=FALSE}
# set working directory to access data
setwd("C:\\Users\\Emilia\\Documents\\Uni Helsinki\\Year One\\Teaching Period 2\\Open Data Science\\IODS-project_2022\\data")

# read data from previously wrangled file 
students_learning2014 <- read.csv("learning_2014.csv", sep = ",", header = TRUE)
```

```{r}
dim(students_learning2014)
str(students_learning2014)
```

*Explanations*

This data set consists of information on 7 variables (X is merely an ID
variable) collected on 166 students. Data was collected from 2014-2015
and the variables selected for this dataset focus on various ways of
learning employed by students, their attitude towards statistics, as
well as background information.

Below you can find a description of each variable present in the
dataset:

| Variable Name | Type             | Descriptor                                       |
|----------------|----------------|----------------------------------------|
| X             | Integer          | order of the variables (1-166)                   |
| Gender        | Character        | student's gender, either M or F                  |
| Age           | Integer          | student's age                                    |
| Attitude      | Numeric (double) | average of student's attitude towards statistics |
| Deep          | Numeric (double) | student's use of deep learning                   |
| Strat         | Numeric (double) | student's use of strategic learning              |
| Surf          | Numeric (double) | student's use of surface learning                |
| Points        | Integer          | student's exam score                             |

### Step 2: Preliminary Exploration

#### 2.1 Graphical Overview

```{r, message = FALSE}
# Graphical Overview 
ggpairs(students_learning2014[-1],                   # data, excluding X
        mapping = aes(col = gender, alpha = 0.3), 
        lower = list(combo = wrap("facethist", bins = 20)))

```

#### 2.2 Summaries

```{r}
students_learning2014[-1]  %>%
  split(.$gender) %>%
  map(summary) #summary(students_learning2014)
```

*Interpretations*

From the graphical representation we can see their frequency,
distributions, scatter plot, and correlations for each variable. Data
from female students is shaded in pink and data from male students is
shaded in light blue. Each variable will be interpreted in more detail
below:

**Gender**

Frequency data reveals that the majority of respondents are female, with
110 respondents being female and 56 being male. There seems to be little
correlation between gender and the other variables. Little is revealed
from the scatter plots.

**Age**

Age is a variable which shows a highly skewed distribution with the
majority of respondents being of a younger age. This is confirmed in the
summary with the mean age being around 25 years old. Younger age seems
to be characteristic irrespective of gender, although it seems that male
respondents have a less sharp distribution curve.

**Attitude**

Attitude possess the highest statistically significant correlation with
the variable points (representing exam scores). As the correlation is
positive and significant (0.437) for both males and females we can
interpret it as the more positive attitude one has towards statistics,
the higher one will perform on exams. It seems that male students have a
slightly more neutral/positive attitude towards statistics as exhibited
in the distribution curve. Female students, having a less sharp
distribution curve, are more neutral/neutral towards statistics with
extreme opinions (points on the tails of the distribution curves) being
infrequent, This is confirmed in the mean attitudes with males having a
mean of 3.443 and females having a mean of 2.990.

**Deep**

Deep has the second highest statistically significant correlation, with
males using surface learning. This is negative correlation meaning that
males who do surface learning are less likely to do deep learning. This
correlation is not significant for female students. Both genders have
similar means for deep learning meaning that males are only slightly
more likely to do deep learning than the other. As the means are higher
than the other learning types (3.657 for females, 3.725 for males), it
means that these students engage in deep learning more often than other
types of learning.

**Strat**

Strategic learning is the second highest type of learning used: in both
males and females (females mean 3.201, males 2.964). Contrasting deep
learning, it appears that female students use strategic learning more
than male students (as seen in the mean). No statistically significant
correlations are present, but the highest correlation is with age. It is
a positive correlation (0.102), meaning that those who are older use
strategic learning more often.

**Surf**

Surface learning has the third highest statistically significant
correlation with the variable attitude. Males are negatively correlated
(-0.374) meaning that male students who do surface learning are more
often have a negative attitude towards statistics. Compared to deep
learning and strategic learning, students appear to use surface learning
less (2.830 mean for females, 2.704 for males).

## Model Building 

#### Step 1: Initial Model 

```{r}
# create and fit regression model with three explanatory variables 
model <- lm(points ~ attitude + age + strat, data = students_learning2014)
summary(model)
```

*Interpretations*

From this model we can see that the combination of the three explanatory
variables can explain up to 20.37% of the variation in a student's exams
scores. While all these variables are significant, only attitude is
significant above the 99% threshold. Age and strategic learning are
significant with a 90% confidence interval with the p-values 0.0981 and
0.0621 respectively.

From the coefficients we can see that for every increase in a positive
attitude the students exam score will increase by 3.48077 points. For
age, as it is negative coefficient, for each increase in age the exam
points will decrease by -0.08822. Finally, for strategic learning, each
increase in use will result in 1.00371 increase in exam points.

This model's F-statistic has a low associated p-value at p= 1.07e-08.

#### Step 2: Re-Fitting the Model

```{r}
model2 <- lm(points ~ attitude, data = students_learning2014)
summary(model2)
```

*Interpretations*

If we remove the explanatory variables that are significant below a 95%
confidence threshold, we are left with the attitude variable. As seen in
the lower adjusted R-squared value (0.1865), this model has lower
predictive power than the previous model. Attitude can explain
approximately 18% of the variation in ones exam scores. An increase in
attitude still results in a slightly higher increase in exam points,
3.5255 compared to 3.48077, and the explanatory is still highly
significant at p = 4.12e-09.

#### Step 3: Model Diagnostics

```{r}
# place all four graphics in same plot
par(mfrow = c(2,2))
# draw diagnostic plots 
plot(model2, which = c(1,2,5))

```

*Interpretations*

From these diagnostics, it can be concluded that the final model is
fairly appropriate. The residuals vs fitted plot shows enough randomness
and does not appear to show and tunnel or curve patterns. The second
plot, normal Q-Q, also confirms the models' appropriateness as these is
little departure from linearity. For the final plot, Residuals vs
Leverage, it looks like no points are outside of Cook's distance meaning
that none of the outliers possess enough leverage or influence to
drastically change the model. However, the points 56 and 35 could be
investigated more closely as they are close.
