# Week 3 - Logistic Regression

This week I have worked with logistic regression. This week was full of many  new concepts for me and I spent many hours on youtube watching "statquest" and "ritvikmath"'s videos on confusion matrices, loss functions and the like. Despite this large learning curve this week was enjoyable and I liked being able to begin to examine the predictive power of models! 

```{r}
date()
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Assignment 3: Logistic Regression

### Relevant Libraries and Packages

```{r, warning=FALSE, message =FALSE}
library(tidyverse)
library(dplyr)
library(ggplot2) 
library(purrr)
library(knitr)
library(patchwork)
library(finalfit)
library(stringr)
library(caTools) 
library(caret)
```

## Step 1: Reading in the Data

```{r, warning=FALSE}
# set working directory to access data
setwd("C:\\Users\\Emilia\\Documents\\Uni Helsinki\\Year One\\Teaching Period 2\\Open Data Science\\IODS-project_2022\\data")

# read data from previously wrangled file 
alc_use <- read.csv("student_alc_use.csv", sep = ",", header = TRUE)
```

```{r}
# explore data structure 
dim(alc_use)
str(alc_use)

# display column names 
print(colnames(alc_use))
```

*Explanations*

Focusing on secondary students' achievement at two schools in Portugal,
this data set contains consisting of 35 variables with 370 observations.
Originally, this data set consisted of two focusing on achievement in
two subjects: math and Portuguese. These two sets were merged by the
author and now contain combined information for each student's
achievement in each subject.

Information about these variables was collected using school reports and
questionnaires the variables included are "student grades, demographic,
social and school related features". From the column names (seen above)
it is clear that this data set covers these four aspects - with a
special focus on social-related features. Variable types vary from
integers, to numerical, character, and logical variables.

Additional attribute information can be found following:
<https://archive.ics.uci.edu/ml/datasets/Student+Performance>

## Step 2: Hypotheses

### Variable 1: Going Out

The first hypothesis to be examined is based on the variable going out
with friends. Essentially, my hypothesis is that those who go out with
their friends more often will be more prone to high alcohol use. I
assume that drinking alcohol is a social bonding activity and this is
why those who go out more frequently are more likely to have higher use.

### Variable 2: Study Time

The second hypothesis is related to the variable study time. My
hypothesis is that those who are studying more each week will have a
lower chance of having high alcohol use. This hypothesis is purely based
on my assumption that if you are spending the majority of your time
studying, you will not have time to develop a high-use alcohol habit.

### Variable 3: Health

The third hypothesis relates to the health variable. As it could
potentially exacerbate existing poor health I hypothesize that those
with poor health are less likely to have high alcohol use.

### Variable 4: Sex

The final hypothesis focuses on the relationship between high alcohol
use and sex. My hypothesis is that male students will have higher rates
of alcohol use.

## Step 3: Numerical and Graphical Exploration

### Numerical

```{r}
keep_columns <- c("high_use",
                  "goout", 
                  "studytime", 
                  "health",
                  "sex"
                  )
alc_small <- select(alc_use, one_of(keep_columns))
```

```{r}
alc_small %>%
  split(.$high_use) %>%
  map(summary)
```

```{r}
dependent   <- "high_use"
explanatory <- c("goout","studytime", "health","sex")

alc_small %>% 
  summary_factorlist(dependent, explanatory, p = TRUE,
                     add_dependent_label = TRUE)
```

*Interpretations*

From the numerical summaries we can see that those with high-use alcohol
consumption are more likely to be male, go out more and spend less time
studying. We can also see from the first summary table that there are
259 students who are not high-use and 111 who are.

It appears that those who go out more often are more likely to have
higher alcohol consumption. This is confirmed in previous summary table
the higher means present in the going out (3.73 for True, 2.853 for
False). These results are significant at a p-value of p = \<0.001.

As for study time, the majority of those with high alcohol use study
below 5 hours a week (42 for under 2 hours and 57 for under 5 hours). We
can also see this in mean summary with a 1.77 average for those with
high-use and a 2.16 average for those who do not have high-use. These
results are significant at p= \<0.001.

With a p value of p = 0.134, the only non-statistically significant
results are those related to health. It can be observed that there is
not a large difference between those who did not drink 3.5 and those who
did 3.7. Perhaps then, in the context of this data set, one's health
does not have a large effect on whether or not one drinks a lot.

Sex, as it is a character, was not covered in the first summary.
However, in the second summary it can be observed that there males
represent 63.1% of the high-use drinkers whereas females represent
36.9%. Rather intuitively these results are flipped with those who are
not high-use. These results are significant at p= \<0.001 which makes it
significant at a 99% CI.

### Graphical

#### Box Plot

```{r}
alc_small %>% 
  ggplot(aes(studytime)) +
  geom_boxplot() +
  scale_y_discrete(breaks = NULL, labels = NULL, name = NULL)
```

*Interpretations*

It appears as though the majority of students study between 1-5 hours a
week (range of values 1 and 2). As for those who study above ten hours a
week (value 4) it appears that these students are outliers in this
dataset.

```{r}
alc_small %>% 
  ggplot(aes(goout)) +
  geom_boxplot() +
  scale_y_discrete(breaks = NULL, labels = NULL, name = NULL)
```

*Interpretations*

Compared to the study time variable, it appears that the going out
variable is normally distributed. The majority of students go out with
their friends an average amount with the tails (very low and very high)
following a normal bell curve.

```{r}
alc_small %>% 
  ggplot(aes(health)) +
  geom_boxplot() +
  scale_y_discrete(breaks = NULL, labels = NULL, name = NULL)

```

*Interpretations*

It appears that the distribution of the health variable is skewed to an
average to very good health status (values 3-5). The majority of
students seem to have reported having a good health status (value 4),
with very few reporting very bad health.

#### Count and Proportions

```{r}
p1 <- alc_small %>% 
  ggplot(aes(x = goout, 
             fill = high_use)) + 
  geom_bar() + 
  theme(legend.position="none")+
  coord_flip() +                               # flip for axis readability + 
  labs(x = "Frequency of Going Out",           # label axes (flipped for coord-flip)
       y = "Count") 

p2 <- alc_small %>% 
  ggplot(aes(x = goout, 
             fill = high_use,
             )) + 
  geom_bar(position = "fill") +
  coord_flip() +                               # flip for axis readability 
  theme(legend.position = NULL)+         
  labs(x = "",                                 # label axes (flipped for coord-flip)
       y = "Proportion") +  
  guides(fill = guide_legend(reverse = FALSE)) # reverse legend order

p1 + p2
```

*Interpretations*

From the proportion visualization, it is clear that those who are going
out more often are more likely to have higher use of alcohol. A clear
jump is seen between the frequency of 3 (average) and 4 (high). Perhaps
this is because the majority of students answered that they see their
friends an average amount and this large n evened out the results.

These results fall in line with my hypothesis that those students who go
out more are have a higher probability of having higher levels of
alcohol consumption.

```{r}
p3 <- alc_small %>% 
  ggplot(aes(x = studytime, 
             fill = high_use)) + 
  geom_bar() + 
  theme(legend.position="none")+
  coord_flip() +                               # flip for axis readability + 
  labs(x = "Time Spent Studying ",             # label axes (flipped for coord-flip)
       y = "Count") 

p4 <- alc_small %>% 
  ggplot(aes(x = studytime, 
             fill = high_use,
             )) + 
  geom_bar(position = "fill") +
  coord_flip() +                               # flip for axis readability 
  theme(legend.position = NULL)+         
  labs(x = "",                                 # label axes (flipped for coord-flip)
       y = "Proportion") +  
  guides(fill = guide_legend(reverse = FALSE)) # reverse legend order

p3 + p4
```

*Interpretations*

Inverse to the going out variable, it appears that the more time one
spends on one's studies, the less likely one is to have higher levels of
alcohol consumption. However, it should be noted that students who study
more than 10 hours a week are outliers within this dataset and therefore
the proportion is perhaps not representative.

These results align with my hypothesis that students who spent more time
studying are less probable to consume alcohol at high levels.

```{r}
p5 <- alc_small %>% 
  ggplot(aes(x = health, 
             fill = high_use)) + 
  geom_bar() + 
  theme(legend.position="none")+
  coord_flip() +                               # flip for axis readability + 
  labs(x = "Health Status",                    # label axes (flipped for coord-flip)
       y = "Count") 

p6 <- alc_small %>% 
  ggplot(aes(x = health, 
             fill = high_use,
             )) + 
  geom_bar(position = "fill") +
  coord_flip() +                               # flip for axis readability 
  theme(legend.position = NULL)+         
  labs(x = "",                                 # label axes (flipped for coord-flip)
       y = "Proportion") +  
  guides(fill = guide_legend(reverse = FALSE)) # reverse legend order

p5 + p6
```

*Interpretations*

Compared to the other two variables, health shows no trend when it comes
to the relationship between health status and alcohol consumption. This
makes sense as it was the only variable which did not possess a
significant p-value (p=0.134).

These results do not fit my hypothesis as I thought that those with
worse health (1 = very bad) would drink less than those with good health
(5 = very good). Therefore, this hypothesis will be rejected.

#### Relationship

```{r}
p1 <- alc_small%>% 
    ggplot(aes(x = high_use, 
               y =goout, 
               col = sex)) +
    geom_point(shape = 1, alpha = 0.5) + 
    geom_jitter(width = 0.1, height = 0.1, alpha = 0.7) +
    labs(x = "High Use of Alcohol", 
         y = "Rate of Going out")

p2 <- alc_small%>% 
    ggplot(aes(x = high_use, 
               y = studytime, 
               col = sex)) +
    geom_point(shape = 1, alpha = 0.5) + 
    geom_jitter(width = 0.1, height = 0.1, alpha = 0.7) +
    labs(x = "High Use of Alcohol", 
         y = "Time Spent Studying")
p3 <- alc_small%>% 
      ggplot(aes(x = high_use, 
                 y = health, 
                 col = sex)) +
      geom_jitter(width = 0.1, height = 0.1, alpha = 0.7) +
      labs(x = "High Use of Alcohol", 
           y = "Current Health Condition")

p1 
p2 
p3
```

*Interpretations*

Like the above visualizations, this one confirms the hypotheses
regarding going out and study time, but rejects the third related to
health. These hypotheses are confirmed through observing the clustering
of points. As this visualization adds the sex dimension, it is
interesting to see if any observations regarding the relationship
between alcohol use and sex can be made.

From the rate of going out visualization it appears to be that those who
have high-use and go out very often are largely male. This can also be
observed in the group of those who go out quite often. For the second
visualization, it appears that those who study less than two hours a
week and have a high-use are more likely to be male. Finally, for the
third visualization no clear patterns in sex are present.

Due to over-fitting it is quite hard to see from this visualization what
role sex might play in high alcohol use. Therefore, I will move on to
the logistic regression section of this report to examine this
hypothesis in more detail.

## Step 4: Logistic Regression

### Model Creation

```{r}
model1 <- glm(high_use ~ goout + studytime + health + sex, data = alc_small, family = "binomial")
```

### Model Summary

```{r}
summary(model1)
```

*Interpretations*

From this model we can see that three of the four explanatory variables
are statistically significant. These three variables are significant at
different confidence intervals however, and only going out is
significant at the 99% threshold. The other two variables, study time
and sex, are significant with a 95% confidence interval (p =3.03e-10 and
a 90% CI (p =0.01073) respectively.

From the coefficients we can see that the effect of going out on high
alcohol use is an increase in the risk for high-alcohol use with the
more on goes out (0,75535). For study time, as it is negative
coefficient, for each increase in time spent studying there will be a
reduction in the risk for high alcohol use (-0.49173). As for health,
there is no significant effect. Finally, for sex, if the sex is male it
will result in a risk increase (0.67717).

### Coefficients and Odds Ratios

```{r}
# compute odds ratios (OR)
OR <- coef(model1) %>% exp

# compute confidence intervals (CI)
CI <- confint(model1) %>% exp 

# print out the odds ratios with their confidence intervals
cbind(OR, CI)
```

*Interpretations*

From the odds ratios we can that those who go out more often are 2.13
times more likely to have high alcohol use. It also is seen that those
of the male sex are 1.97 times more likely to have high alcohol use
level. Comparatively, study time has a lower ratio with it being only
0.61 times more likely to have high alcohol use if you have

Below, the odds ratio and the confidence interval difference can be
seen. The confidence interval difference represents the range which is
represented within the 95% confidence interval.

| Variable   | OR   | Confidence Interval Difference |
|------------|------|--------------------------------|
| Going out  | 2.13 | 1.02 (1.01951459)              |
| Study Time | 0.61 | 0.41 (0.4120806)               |
| Health     | 1.11 | 0.40 (0.40074558)              |
| Sex        | 1.97 | 2.15 (2.15400134)              |

## Step 5: Predictive power

Using the variables which, according to your logistic regression model,
had a statistical relationship with high/low alcohol consumption,
explore the predictive power of you model.

#### Model Creation

```{r}
final_model <- glm(high_use ~ goout + studytime + sex, data = alc_small, family = "binomial")
```

Provide a 2x2 cross tabulation of predictions versus the actual values
and optionally display a graphic visualizing both the actual values and
the predictions.

#### Cross tabulations: Predicted vs Actual

##### Numerical

```{r}
# predict() the probability of high_use
probabilities <- predict(final_model, type = "response")

# add the predicted probabilities to 'alc_small'
alc_small <- mutate(alc_small, probability = probabilities)

# use the probabilities to make a prediction of high_use
alc_small <- mutate(alc_small, prediction = alc_small$probability > 0.5)

# tabulate the target variable versus the predictions
table(high_use = alc_small$high_use, prediction = alc_small$prediction)
```

##### Total Training Error 

proportion = `(false positive +  false negative)/n`

proportion = `(25 + 58)/370`

proportion = 0.2243

```{r}
confusionMatrix(
  factor(alc_small$prediction), 
  factor(alc_small$high_use))
```

*Interpretations*

As we can see from the confusion matrix and total training error
proportion, most of the time the model is right but it is wrong enough
of the time to not be helpful. We can see that the total training error is around 22%, meaning that the model is inaccurate 22% of the time and accurate 77%. The model struggles to predict true positives as accurately and produces a lot of false positives. I am not entirely sure
why this is, perhaps because the data the model was trained on did not contain enough when positive high-use values when compared to false values.

Although I am not at the point where I can interpret the other outputs
(Kappa, Sensitivity etc.), I hope to reach there one day!

##### Graphical

```{r}
# create confusion matrix 
cmat <- confusionMatrix(
  factor(alc_small$prediction), 
  factor(alc_small$high_use))

# visualize confusion matrix 
fourfoldplot(cmat$table, color = c("cyan", "pink"),
             conf.level = 0, margin = 1, main = "Confusion Matrix")

```

*Final Interpretations and Model/Hypotheses Comparison*

A generous reading of the confusion matrix and the performance of the
model would say that the model predicts false values correctly 9/10 of
the time and positive values correctly 1/2 of the time. I do not believe
that my simple guessing strategies would be able to reach this level of
accuracy. When guessing, I often rely on broad generalizations (those
who go out must drink more) to reach my conclusions. My hypotheses were
thus oversimplified and did not reveal the nuance in the data. A clear
example was that those who go out often have a 50% chance of having high
alcohol use or not. Therefore, in order to reach the model accuracy, I
would need to take more factors into account during my guessing. Yes, it
is true that male students do drink more but is how much they drink
proportionally to their weight taken into account? Therefore, while a
simple guessing strategy can perhaps point the researcher which way to
look, I do not think it can predict results as accurately as a
well-built model can.
